Reasonably Programmable Literal Notation

### 2018-03-07 icfp18/syntax-icfp18.tex
revisions to page 5

Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}. Regexes are  common in, for example, bioinformatics, where they are used to express patterns in DNA sequences. For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
\begin{lstlisting}[numbers=none]
  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
\end{lstlisting}
The notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify each application of \li{Regex.Or} and \li{Regex.Str}. Even with this shorthand, however, this notation for constructing regexes is rather unwieldy. Instead, we would like to be able to use the common POSIX-style notation for constructing values of type \li{Regex.t} \cite{STD95954}.

@@

\vspace{-3px}
\subsection{TLM Definition and Application}
Figure \ref{fig:regex-tlm-def} defines a TLM named \li{$regex} for values of type \li{Regex.t} that supports a POSIX-style literal notation extended with splice forms for regexes and strings. Line 1 of Fig. \ref{fig:first-tlm-example} applies this TLM to construct the regex \li{any_base} just described. Line 2 of Fig. \ref{fig:first-tlm-example} then applies \li{$regex} again, using its regex splice form to construct a regex \li{bisA} matching the DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base. Finally, Lines 3-4 of Fig. \ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string, which is spliced in using a string splice form delimited by \li{$$(} and \li{)}.

Let us consider the second of these three TLM applications more closely:
\begin{lstlisting}[numbers=none]
    $regex `(GC$(any_base)GC)`
\end{lstlisting}
According to the context-free grammar of the language, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
 to distinguish them from variables. There are no lexical constraints on the body of the generalized literal form except that any occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in OCaml, Reason and other languages. Generalized literal forms, which first arose in the prior work on TSLs \cite{TSLs}, therefore syntactically subsume other literal forms. The prior work specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use \li{`(} and \li{)`} exclusively. The fact that the context-free grammar of the base language is fixed ensures that notation providers can reason modularly about syntactic ambiguity in their own grammars.

@@

% \section{Simple Expression TLMs By Example}\label{sec:tsms-by-example}
% We begin in this section with a ``tutorial-style'' introduction to seTLMs in Reason. %In particular, we will define an seTLM for constructing values of the recursive labeled sum type \li{rx} that was defined in Figure \ref{fig:datatype-rx}. 
% Sec. \ref{sec:tsms-minimal-formalism} then formally defines a reduced dialect of Reason called $\miniVerseUE$. This will serve as a ``conceptually minimal'' core calculus of TLMs, in the style of the simply typed lambda calculus.   %We conclude in Sec. \ref{sec:uetsms-discussion}

>>>>

Consider the recursive datatype \li{Regex.t} defined by the module in Fig.~\ref{fig:Regex-module-def}, which encodes {regular expressions} (regexes) into Reason \cite{Thompson:1968:PTR:363347.363387}. Regexes are  common in, for instance, bioinformatics, where they are used to express patterns in DNA sequences. For example, we can construct a regex that matches the strings \li{"A"}, \li{"T"}, \li{"G"} or \li{"C"}, which represent the four bases in DNA, as follows:
\begin{lstlisting}[numbers=none]
  let any_base = Regex.(Or(Str "A", Or(Str "T", Or(Str "G", Or(Str "C")))))
\end{lstlisting}
The notation \li{Regex.(}$e$\li{)} locally opens the module \li{Regex} for use within the expression $e$, so we do not need to qualify each application of \li{Regex.Or} and \li{Regex.Str}. Even with this shorthand, however, this notation for constructing regexes is rather unwieldy. Instead, we would like to be able to use the common POSIX-style notation  \cite{STD95954} for constructing values of type \li{Regex.t}, including values constructed compositionally from previously constructed regexes and strings.

@@

\vspace{-3px}
\subsection{TLM Definition and Application}
We solve this problem in Figure \ref{fig:regex-tlm-def} by defining a TLM named \li{$regex} (pronounced ``lit regex'') that supports POSIX-style regex notation extended with splice forms for regexes and strings. We will consider the provider perspective, i.e. the details of \li{RegexLexer} and \li{RegexParser}, in Sec. \ref{sec:provider-perspective}, but the focus of this paper is on the client perspective, with an emphasis on holding the details of the lexer and parser abstract, so let us start instead with some examples of \li{$regex} being applied.

Line 1 of Fig. \ref{fig:first-tlm-example} applies \li{$regex} to construct the regex \li{any_base} just described. Line 2 of Fig. \ref{fig:first-tlm-example} then applies \li{$regex} again, using its regex splice form to construct a regex \li{bisA} matching the DNA sequences recognized by the BisA restriction enzyme, where the middle base can be any base. Finally, Lines~3-4 of Fig. \ref{fig:first-tlm-example} define a function that constructs a more complex regex from these two regexes and a given gene sequence represented as a string, which is spliced in using the string splice form. Notice that the TLM provider has chosen to use \li{$(}$e$\li{)} for regex splicing and \li{$$(}$e$\li{)} for string splicing, where $e$ is an arbitrary Reason expression.

Let us consider the second of these three TLM applications more closely:
\begin{lstlisting}[numbers=none]
    $regex `(GC$(any_base)GC)`
\end{lstlisting}
According to the context-free grammar of (this paper's dialect of) Reason, this form is simply a leaf of the unexpanded syntax tree, like a string literal would be. TLM names are prefixed by \li{$} 
 to distinguish them from variables. We call \li{`(GC$(any_base)GC)`} a generalized literal form, following the terminology introduced in the prior work on type-specific languages \cite{TSLs}. There are no lexical constraints on the body of a generalized literal form except that any occurrences of \li{`(} must be balanced by \li{)`}, much like nested comments in Reason/OCaml. Generalized literal forms therefore syntactically subsume other literal forms. The nesting constraint is to support TLM applications that appear inside spliced expressions (discussed below). The prior work specified several other choices of outer delimitation, including layout-sensitive delimitation, but for the purposes of this work, we will use \li{`(} and \li{)`} exclusively.

@@

\subsection{Provider Perspective}
\label{sec:provider-perspective}

The fact that the context-free grammar of the base language is fixed ensures that notation providers can reason modularly about syntactic ambiguity in their own grammars.
% \section{Simple Expression TLMs By Example}\label{sec:tsms-by-example}
% We begin in this section with a ``tutorial-style'' introduction to seTLMs in Reason. %In particular, we will define an seTLM for constructing values of the recursive labeled sum type \li{rx} that was defined in Figure \ref{fig:datatype-rx}. 
% Sec. \ref{sec:tsms-minimal-formalism} then formally defines a reduced dialect of Reason called $\miniVerseUE$. This will serve as a ``conceptually minimal'' core calculus of TLMs, in the style of the simply typed lambda calculus.   %We conclude in Sec. \ref{sec:uetsms-discussion}
